## Small Samples {#sec-energy-intro .unnumbered}




```{r}
#| warning: false
#| eval: true
#| echo: false
#| message: false

library(here)
invisible(capture.output(source(here("scripts", "global.R"))))
invisible(capture.output(source(here("scripts", "helpers.R"))))
invisible(capture.output(source(here("scripts", "gamma_funcs.R"))))



```


::: {.callout-tip title="Aggregation heals mean, not Variance"}

In practice, we have only few years of observation (1-5), which is equivalent to our Monte Carlo with low $n_{sim}$. Our observed data $Y$, however, is composed of a large number of aggregated policies $k$ (certainly >100).
We learn that in our setting with small $n_{sim}$ that large $k$ does heal mean, but not variance. In our simulation with $n_{sim}=2$ (2 years), the sample mean quickly convergence to the true mean as $k>20$, but the $SD$ remains completely off. In fact, the variance gets very tight around the (correct) sample mean and completely misses the tail.

#### Critical
Large $k$ cures the mean via the law of large numbers inside each year, but it cannot cure the tail when $n_{sim}$ (years) is small. Intuitively, with $n≤5$ the chance to observe an extreme event (year) is small, no matter how many policies are aggregated. The LLN gives us false confidence and we might drastically underestimate the tail.

:::

\
\



### Monte Carlo Simulations with small samples

Below we now vary the number of MS simulations and observe the effect it has on the estimated mean and variance.

Note that in our risk model, the $n_{sim}$ corresponds to the number of yearly observations we have for one policy, whereas $k$ signifies the number of policies in an aggregated, observed municipality data point. Our goal here is to understand the effects of both $n_{sim}$ and $k$.


```{r, fig.height=10, fig.width=10}
#| warning: false
#| eval: true
#| echo: false
#| message: false

make_gamma_mc_plots(theta = c(0.1),  alpha = c(1.3), k = c(1), n_sims = c(2, 5, 20, 50, 500, 5000) )
```


```{r, fig.height=10, fig.width=10}
#| warning: false
#| eval: true
#| echo: false
#| message: false
make_gamma_mc_plots(theta = c(0.1),  alpha = c(1.3), k = c(3), n_sims = c(2, 5, 20, 50, 500, 5000) )

```

```{r, fig.height=10, fig.width=10}
#| warning: false
#| eval: true
#| echo: false
#| message: false

make_gamma_mc_plots(theta = c(0.1),  alpha = c(1.3), k = c(9), n_sims = c(2, 5, 20, 50, 500, 5000) )

```


```{r, fig.height=10, fig.width=10}
#| warning: false
#| eval: true
#| echo: false
#| message: false

make_gamma_mc_plots(theta = c(0.1),  alpha = c(1.3), k = c(27), n_sims = c(2, 5, 20, 50, 500, 5000) )
```


```{r, fig.height=10, fig.width=10}
#| warning: false
#| eval: true
#| echo: false
#| message: false

make_gamma_mc_plots(theta = c(0.1),  alpha = c(1.3), k = c(100), n_sims = c(2, 5, 20, 50, 500, 5000) )
```



### Aggregation heals mean, but not variance

In practice, we have only few years of observation (1-5), which is equivalent to our monte carlo with few draws. Our data, however, is composed of a large number of aggregated fields (certainly >100), i.e. k is large. It's important to note that the aggregation effect heals the mean (given i.i.d.), but NOT the variance.

```{r, fig.height=10, fig.width=10}
#| warning: false
#| eval: true
#| echo: false
#| message: false

make_gamma_mc_plots(theta = c(0.1),  alpha = c(1.3), k = c(1, 100), n_sims = c(2, 500) )
```



Worse even... as we increase k, sd gets smaller, but likely wrong.**Hence we give ourselves a wrong confidence**

```{r, fig.height=10, fig.width=10}
#| warning: false
#| eval: true
#| echo: false
#| message: false

make_gamma_mc_plots(theta = c(0.1),  alpha = c(1.3), k = c(1, 3, 9, 27), n_sims = c(2) )

```


